Query Optimization for Dynamic Imputation

### 2017-03-01 paper/experiments.tex

We also calculate SMAPE over the number of tuples used to produce the results
in each configuration with respect to the number of tuples satisfying the same query when
the base tables have been imputed. In many cases where the SMAPE reduction between $\alpha = 1$ and
$\alpha = 0$ is small, for example \Cref{q6}, we highlight that the count of tuples can be significantly different between
the two scenarios. So while the aggregate used might not have been significantly affected by the different population,
other aggregates may well be. This highlights a challenge for a user handling data imputation traditionally: it is unclear if
the missing data will have a large or small negative impact on their analysis until they have paid the cost of
running it. By using \ProjectName{} this cost can be lowered significantly.

\begin{table}
\centering
\input{tables/perf_summary_13.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations. To calculate SMAPE, \ProjectName{} results are compared to query results returned
    from executions with base table imputation. Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). Count shares show the number of results returned from each
    query as a percentage of the number of results returned when imputing on the base table.
    A lower count share reflects more potential for errors.}
\label{table:smape}
\end{table}

>>>>

We also calculate SMAPE over the number of tuples used to produce the results
in each configuration with respect to the number of tuples satisfying the same query when
the base tables have been imputed. In many cases where the SMAPE reduction between $\alpha = 1$ and
$\alpha = 0$ is small, for example \Cref{q6}\todo{confirm}, we highlight that the count of tuples can be significantly different between
the two scenarios. So while the aggregate used might not have been significantly affected by the different population,
other aggregates may well be. This highlights a challenge for a user handling data imputation traditionally: it is unclear if
the missing data will have a large or small negative impact on their analysis until they have paid the cost of
running it. By using \ProjectName{} this cost can be lowered significantly.

\begin{table}
\centering
\input{tables/perf_summary.tex}
\caption{Symmetric-Mean-Absolute-Percentage-Error for queries run under different $\alpha$
    parameterizations. To calculate SMAPE, \ProjectName{} results are compared to query results returned
    from executions with base table imputation. Queries optimized
    for quality ($\alpha=0$) generally achieve lower error than queries optimized for
    efficiency ($\alpha=1$). Count shares show the number of results returned from each
    query as a percentage of the number of results returned when imputing on the base table.
    A lower count share reflects more potential for errors.}
\label{table:smape}
\end{table}

### 2017-03-01 paper/introduction.tex

A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows.  Also unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on the  query planning involved in optimal placement of imputation operations
in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  The drop operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an additional imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

>>>>

A simpler approach might drop all rows with missing values, which can not only introduce bias into the results, but also result in discarding all of the data.
In contrast to existing systems~\cite{burgette2010multiple,akande2015empirical}, \ProjectName{} avoids imputing over the entire data set, as done with traditional methods. Specifically, \ProjectName{}
carefully plans the placement of imputation / row drop operations inside the query plan, resulting in a 
significant speedup in query execution while minimizing the number of dropped rows. Unlike previous work, the focus on of our work is not on the imputation algorithms themselves (we can employ almost any such algorithm), but rather on placing imputation operations
optimally in query plans.   Specifically, our optimization algorithms generate the Pareto-optimal frontier of plans that trade imputation cost for result quality.

@@

\subsection{Contributions}
\ProjectName{} leverages a number of contributions to minimize imputation time while producing comparable results to traditional approaches,
increasing productivity for users who frequently encounter missing values in their data. These contributions include:
\begin{itemize}
\item \textbf{Relational algebra extended with imputation:}
  We extend the standard relational algebra with two new operators to represent imputation operations: impute ($\mu$) and drop ($\delta$).
  The impute operation fills in missing data values using an iterative regression model (CE-CART)~\cite{burgette2010multiple}.
  The drop operation simply drops tuples which contain \nullv{} values.
\item \textbf{Model of imputation quality and cost:}
  We extend the traditional cost model for query plans to incorporate a measure of the quality of the imputations performed.
  We use the cost model to abstract over the imputation algorithm used.
  To add an imputation technique, it is sufficient to characterize it with two functions: one to describe its running time and one to describe the quality of its results.
\item \textbf{Query planning with imputation:}
  We present the first query planning algorithm to jointly optimize for running time and the quality of the imputed results.
  It does so by maintaining multiple sets of Pareto optimal plans according to the cost model.
  By deferring selection of the final plan, we make it possible for the user to trade off running time and result quality.
%\item \textbf{Proof of optimality:}
% We prove that our cost-model and planning algorithm produce a sound and complete final Pareto frontier, making the final plan choice optimal under our search space and the user's preferences.
\end{itemize}

### 2017-03-01 paper/relatedwork.tex

Imputation of missing values is widely studied in the statistics and machine
learning communities. As highlighted in~\cite{gelman2006data}, missing data
can appear for a variety of reasons.
It can be missing uniformly at random and or conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between attributes to account for different forms of missingness. For example, Burgette and Reiter~\cite{burgette2010multiple} discuss the usage of sequential regression trees
for imputing missing data.

In~\cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well-known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a trade-off between
imputation quality and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

\ProjectName's cost-based query planner 
is partially based on the seminal work developed for System R's query planning~\cite{blasgen1981system}.
However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
uses histogram transformations to account for the changing nature of missing values.

>>>>

Imputation of missing values is widely studied in the statistics and machine
learning communities. As highlighted in~\cite{gelman2006data}, missing data
can appear for a variety of reasons.
It can be missing uniformly at random or conditioned on
existing values (observed and missing). Methods in the statistical community
focus on correctly modeling relationships between attributes to account for different forms of missingness. For example, Burgette and Reiter~\cite{burgette2010multiple} discuss the usage of sequential regression trees
for imputing missing data.

~\cite{akande2015empirical} analyzes the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well-known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a trade-off between
imputation quality and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

\ProjectName's cost-based query planner 
is partially based on the seminal work developed for System R's query planning~\cite{blasgen1981system}.
However, in contrast to System R, \ProjectName{} performs additional optimizations for imputing missing data and
uses histogram transformations to account for the way relational and imputation operators affect the underlying
tuple distributions.
